Towards Verifying the Bitcoin-S Library

### 2019-06-27 paper/paper.tex
more work on transformations

\lstinputlisting[style=scala]{../code/addition/src/main/scala/addition/reduced/currency/CurrencyUnits.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/addition/reduced/number/NumberType.scala}

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:33:1: Objects cannot extend classes or implement traits, use a case object instead
          object Satoshis extends BaseNumbers[Satoshis] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[ Error  ] number/NumberType.scala:65:1: Objects cannot extend classes or implement traits, use a case object instead
          object Int64 extends BaseNumbers[Int64] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
\end{lstlisting}

Here, we can just turn the objects into case objects by literally just
changing the word \texttt{object} into \texttt{case object} on lines
33 and 65 in the two respective files.

@@

This should be easy to rewrite by using generics instead of an abstract type, right?
Unfortunately not.
The problem is, CurrencyUnit uses one of its implementing classes: Satoshis.

Before:
\begin{lstlisting}[style=scala]
sealed abstract class CurrencyUnit {
  type A

  def satoshis: Satoshis

  protected def underlying: A
}

sealed abstract class Satoshis extends CurrencyUnit {
  override type A = Int64

  override def satoshis: Satoshis = this

  def toBigInt: BigInt = BigInt(toLong)

What happens, if we typify CurrencyUnit with A, meaning to make it generic with type A?

Satoshis extends CurrencyUnit with type Int64, so it would be of type CurrencyUnit[Int64].
That's too specific, because the return type of the addition is then CurrencyUnit[Int64] not CurrencyUnit[A].

Maybe the Bitcoin-S developers should reimplement this part and not use Satoshis directly.
\todo{ramon: explain. what would be the correct code?}

@@

After:
\begin{lstlisting}[style=scala]
sealed abstract class CurrencyUnit {
  def satoshis: Satoshis

  protected def underlying: Int64
}

sealed abstract class Satoshis extends CurrencyUnit {
  override def satoshis: Satoshis = this

  def toBigInt: BigInt = BigInt(toLong)

@@

Stainless output:
\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:26:33: Only literal arguments are allowed for BigInt.
            def toBigInt: BigInt = BigInt(toLong)
                                          ^^^^^^
\end{lstlisting}

As described before, Stainless supports only a subset of Scala.
The BigInt from the Stainless library is a bit restricted.
One such restriction is, that BigInt does not support dynamic BigInt construction.
Thus, the constructor parameter of BitInt must be a literal argument.

  def toBigInt: BigInt = BigInt(toLong)

@@

This would be really hard to refactor, because Bitcoin-S tries to be as dynamic as possible so it can be used with cryptocurrencies other than bitcoins.
Maybe it could be impossible, because they need to parse dynamic values from the bitcoin network.

Luckily, we can use toBigInt on the field \texttt{underlying} directly instead of toLong.
So, instead of converting the underlying to Long and back to BigInt we convert underlying directly to BigInt.

After fixing all Stainless errors, a new error appears.

@@

Stainless output:
\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:8:30: Unknown type parameter type T
          sealed abstract class Number[T <: Number[T]] {
                                        ^^^^^^^^^^^^^^
\end{lstlisting}

This is a missing feature in Stainless.
It does not support upper type boundaries on the class itself.
To track this, \href{https://github.com/epfl-lara/stainless/issues/519}{issue 519} was created on GitHub during this work.

Before:
\begin{lstlisting}[style=scala]
sealed abstract class Number[T <: Number[T]] {
  type A = BigInt

  /** Factory function to create the underlying T, for instance a UInt32 */
  def apply: A => T

  def +(num: T): T = apply(checkResult(underlying + num.underlying))

/**
  * Represents a signed number in our number system
  * Instances of this is [[Int64]]
  */
sealed abstract class SignedNumber[T <: Number[T]] extends Number[T]

/**
  * Represents a int64_t in C
  */
sealed abstract class Int64 extends SignedNumber[Int64] {
  override def apply: A => Int64 = Int64(_)
  override def andMask = 0xffffffffffffffffL
}
\end{lstlisting}

Despite this, in order to be able to continue, we make this a concrete type by replacing T with Int64.
Int64,  because Satoshis uses only Int64.
There are other number types like UInt16 but for our property we don't need them.

Now, there are two new errors.

\subsection{Missing Member \texttt{bigInteger} in BigInt}
\todo{Ramon is .bigInteger missing in Stainless?}
Stainless output:
\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:14:22: Unknown call to bigInteger on Number.this.toBigInt (BigInt) with arguments List() of type List()
            def toLong: Long = toBigInt.bigInteger.longValueExact()
                              ^^^^^^^^^^^^^^^^^^^
\end{lstlisting}

>>>>

\lstinputlisting[style=scala]{../code/addition/src/main/scala/addition/reduced/number/NumberType.scala}
\lstinputlisting[style=scala]{../code/addition/src/main/scala/addition/reduced/currency/CurrencyUnits.scala}

\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:65:1: Objects cannot extend classes or implement traits, use a case object instead
          object Int64 extends BaseNumbers[Int64] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
[ Error  ] currency/CurrencyUnits.scala:33:1: Objects cannot extend classes or implement traits, use a case object instead
          object Satoshis extends BaseNumbers[Satoshis] {
          ^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^...
\end{lstlisting}

Here, we can just turn the objects into case objects by literally just
changing the word \texttt{object} into \texttt{case object} on lines
65 and 33 in the two respective files.

@@

Note that we can not simply replace the unsupported abstract type
member by a (supported) type parameter. The problem is that the
CurrencyUnit class uses one of its implementing classes: Satoshis.

Satoshis would have to instantiate a potential type parameter with
type Int64, so it would extend CurrencyUnit[Int64]. But that is too
specific, because the return type of the \texttt{+}-method would then then be
CurrencyUnit[Int64] not CurrencyUnit[A].

@@

We remove line 6 and line 22 from CurrencyUnits.scala (to maintain
line numbers we actually replace them with empty lines for now) and in
line 18 we replace A by Int64.

@@

\begin{lstlisting}[style=stainless]
[ Error  ] currency/CurrencyUnits.scala:26:33: Only literal arguments are allowed for BigInt.
            def toBigInt: BigInt = BigInt(toLong)
                                          ^^^^^^
\end{lstlisting}

As described before, the types in the Stainless library are more
restricted than their Scala library counterparts. In particular, the
Stainless BigInt constructor is restricted to literal arguments.

@@

Here we can simply use toBigInt on the field \texttt{underlying}
directly.  So, instead of converting the underlying to Long and back
to BigInt we convert underlying directly to BigInt.

We replace line 26 by
\begin{lstlisting}[style=scala]
def toBigInt: BigInt = underlying.toBigint
\end{lstlisting}

\todo{explain why equivalence or why not equivalence}

@@

\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:8:30: Unknown type parameter type T
          sealed abstract class Number[T <: Number[T]] {
                                        ^^^^^^^^^^^^^^
\end{lstlisting}

Stainless does not currently support a class with a type parameter
with a type boundary that contains the class itself. We opened an
issue \cite{Stainless:issue519} on the Stainless repository and the
Stainless developers have targeted Stainless version 0.4 to support
such self-referential type boundaries.

For now, since our code only uses Number with type parameter T
instantiated to Int64, we just remove the type parameter and replace
it by Int64. We respectively replace lines 8, 44 and 49 by
\begin{lstlisting}[style=scala]
sealed abstract class Number {
\end{lstlisting}

\begin{lstlisting}[style=scala]
sealed abstract class SignedNumber extends Number
\end{lstlisting}

\begin{lstlisting}[style=scala]
sealed abstract class Int64 extends SignedNumber {
\end{lstlisting}

and replace T by Int64 in lines 25-27.

\subsection{Missing Member \texttt{bigInteger} in BigInt}

\todo{Ramon is .bigInteger missing in Stainless?}

Stainless output:
\begin{lstlisting}[style=stainless]
[ Error  ] number/NumberType.scala:14:22: Unknown call to bigInteger on Number.this.toBigInt (BigInt) with arguments List() of type List()
            def toLong: Long = toBigInt.bigInteger.longValueExact()
                              ^^^^^^^^^^^^^^^^^^^
\end{lstlisting}
